User Guide for Corrfunc

### 2015-01-28 paper/users-guide.tex
Latest version. Made some changes to tests suite in case of failures

\documentclass[12pt,titlepage,justified]{article}

\usepackage{amsmath}
\usepackage{aas_macros}
\usepackage{rotating}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{dirtree}
\usepackage{adjustbox}
\usepackage{bytefield}
\usepackage{chngcntr}
\usepackage{alltt}

@@

%% Code paper - so there are lots of code listings
\lstset{
  language=C,                % choose the language of the code
  basicstyle=\scriptsize\ttfamily,      % sets font
  numbers=none,                   % where to put the line-numbers
%%   numberstyle=\small, 
%%   stepnumber=1,                   % the step between two line-numbers.
%%   numbersep=10pt,                  % how far the line-numbers are from the code
  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  tabsize=2,                      % sets default tabsize to 2 spaces
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=true,         % sets if automatic breaks should only happen at whitespace
  captionpos=t,  
  xleftmargin=17pt,
  framexleftmargin=17pt,
  framexrightmargin=17pt,
  framexbottommargin=5pt,
  framextopmargin=5pt,
  showstringspaces=false,
  frame=top,frame=bottom,
  escapeinside={(*@}{@*)},
}
%% Change the listings counter to show Section number + listings number (rather than just the listings number)
\AtBeginDocument{\counterwithin{lstlisting}{section}}

@@

Cache locality and hand-written AVX intrinsics are the reasons why the code is very fast. However, that also means that the code is 
not very portable. I have tried my best to ensure that the codes work on Linux and MAC OSX. If it does not work for you, particularly 
if you are on a reasonable Linux install, please \href{mailto:manodeep@gmail.com}{email me}. The API for the codes can be considered 
frozen; I will not change the API without changing the MAJOR release version.

The paper associated with the codes is being (slowly) written by me. I hope to submit the paper to Astronomy \& Computing and release 
the codes simultaneously. Once I submit the paper to arXiv, there will be an actual paper arXiv link \href{http://adsabs.harvard.edu/abs/2014arXiv1405.5832L}{here}.

@@

\subsection{Getting the Source}
You can obtain the source in two ways: i) Clone the mercurial repo (\texttt{hg clone} \url{https://bitbucket.org/manodeep/corrfunc/}) 
or ii) Download the tar archive (\href{https://bitbucket.org/manodeep/corrfunc/downloads/corrfunc.1.0.0.tar.gz}{corrfunc.\$MAJOR.0.\$MINOR.tar.gz}) and 
unpack it in the directory where you wish to keep the files (\texttt{tar xvzf corrfunc.\$MAJOR.0.\$MINOR.tar.gz}). Here, \$MAJOR and \$MINOR
refer to the major and minor release versions (current \$MAJOR=1, \$MINOR=0). I will only change the \$MAJOR version if the API breaks.

The directory structure for the code looks like this:
\dirtree{%
.1 corrfunc.
.2 paper.
.2 xi\_theory.
.3 benchmarks\DotFill
  \begin{minipage}[t]{8cm}
    IDL scripts to run benchmarks{.} 
  \end{minipage}.
.3 bin\DotFill 
  \begin{minipage}[t]{8cm}
    Will be created to copy executable files when you run `make install'{.} 
  \end{minipage}.
.3 examples\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for example C bindings using the static libraries{.} 
  \end{minipage}.
.3 include\DotFill 
  \begin{minipage}[t]{8cm}
    Header files for static libraries{.} 
  \end{minipage}.
.3 io\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for reading in data{.} 
  \end{minipage}.
.3 lib\DotFill 
  \begin{minipage}[t]{8cm}
    Will be created to copy static libraries and python library after you run `make'{.} 
  \end{minipage}.
.3 python\_bindings\DotFill 
  \begin{minipage}[t]{8cm}
    Source files to generate python bindings{.} 
  \end{minipage}.
.3 tests\DotFill 
  \begin{minipage}[t]{8cm}
    Correct outputs for tests{.}
  \end{minipage}.
.4 data\DotFill 
  \begin{minipage}[t]{8cm}
    Mock galaxy catalogs for tests{.} 
  \end{minipage}.
.3 utils\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for creating 3-D grid and helper routines{.} 
  \end{minipage}.
.3 xi\_of\_r\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $\xi(r)${.} 
  \end{minipage}.
.3 xi\_rp\_pi\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $\xi(r_p,\pi)${.} 
  \end{minipage}.
.3 wp\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $w_p(r_p)${.} 
  \end{minipage}.
}

\subsection{Code Options}
There are a few code options that control both the Science case and the code compilation. All of these 
options are located in `common.mk' in the base directory (`corrfunc'). Edit the first few lines to set these 
options (see Table.~\ref{table:options} for details) :
\begin{itemize}
\item Science options -- PERIODIC, OUTPUT\_RPAVG
\item Code options -- DOUBLE\_PREC, USE\_AVX and USE\_OMP
\end{itemize}

@@

\flushleft The \texttt{fast-food} file consists of a header:
\begin{lstlisting}[language=C,numbers=none,caption={The header format for \texttt{fast-food} files}]
int idat[5];
float fdat[9];
float znow;
\end{lstlisting}
For the purposes of these correlation function codes, the only useful quantity is \texttt{idat[1]} which contains \texttt{N} -- the number of particles 
in the data file.

@@

\stdsection{Code Design}
The following sections are taken directly from the associated \href{http://adsabs.harvard.edu/abs/2014arXiv1405.5832L}{paper}. The fundamental idea 
behind the code can be broken down into the following steps:
\begin{itemize}
\item Given the max. separation, \rmax, for the correlation function, grid the entire the particle distribution with cell width \rmax. In case of 
\xirppi and \wprp, the \texttt{X/Y} bins are \rmax while the \texttt{Z} bins is \pimax.
\item In each 3-D cell, store all the particles that are located in that cell in contiguous \texttt{X/Y/Z} arrays (see Listing~\ref{code:assign_to_cellarray}). The particle data are contained in a structure 
defined in Listing~\ref{code:celldefn}.
\item Loop over all particles in every cell (this loop uses OpenMP parallelization if USE\_OMP is enabled). This is the \texttt{index1} loop in 
\texttt{countpairs.c, countpairs\_rp\_pi.c and countpairs\_wp.c}. The loop variable for the loop over each particle is \texttt{i}.
\item Now, for every particle in the input list, we only need to look at particles in the neighbouring cells (see Fig.~\ref{fig:grid}). 
\item Loop over neighbouring cells in all 3 dimensions. Corresponding variables are \texttt{iiix, iiiy, iiiz}. These variables are constructed from the 
loop variables \texttt{iix, iiy, iiz} to account for periodic boundary conditions. \footnote{I apologize for this atrocious variable naming scheme.}
\item Once we have a triplet of \texttt{iiix, iiiy, iiiz}, we can construct the index for the neighbouring cell. Variable \texttt{index2} gives the index 
for the neighbouring cell. 
\item Compute the distances between target particle and some particle bunch (where a chunk is 4 double or 8 floats) using AVX intrinsics. The slower, non-AVX 
version of the code computes the distances serially and does not take advantage of the CPU architecture. This is the \texttt{j} loop in 
\texttt{countpairs.c, countpairs\_rp\_pi.c, countpairs\_wp.c}. 
\item Check if any of the distances are less than \rmax. If not, continue processing the data. If yes, then update the histograms using 
AVX \texttt{bit-masks} and \texttt{popcount}. 
\item Continue until all particles in neighbouring cell are done. 
\end{itemize}

@@

\subsection{How to Maintain Cache Locality within the Grid}
For all pairs around a given target galaxy, we need to compute distances to all points within all neighbouring 3-d cells.
We ensure that the particle locations are contiguous by moving them into the following \texttt{C struct} in the order in which they arrive.
\begin{lstlisting}[language=C,numbers=none,label={code:celldefn},caption={Definition of the cellarray structure. This 
structure contains the \texttt{X/Y/Z} positions of all the particles that are in one 3-D cell. }]
typedef struct{
  DOUBLE *x;
  DOUBLE *y;
  DOUBLE *z;
  int64_t nelements;
} cellarray;
\end{lstlisting}
The code \texttt{gridlink.c} takes in an input list of 3 arrays \texttt{X/Y/Z} and grids them into a regular 3-D grid using the specified bins parameter
\texttt{max\_x\_size} or \texttt{max\_y\_size} or \texttt{max\_z\_size} for the \texttt{X/Y/Z} axes respectively. \footnote{In practice, the bins may be further 
subdivided using the corresponding bin refine factors. These bin refine factors seem to influence runtime the most, you should experiment with a few values of 
\texttt{bin\_refine\_factor} and \texttt{zbin\_refine\_factor} to see what produces the best runtimes for your typical scenario.}
Once the number of grid cells along each axes has been determined, we allocate memory for the \texttt{struct lattice}. This \texttt{struct lattice} is a 
declared as an 1-D array; the conversion from the three indices in 3-D (\texttt{ix,iy,iz}) to a single index (\texttt{index}) happens through the last 
line in Listing~\ref{code:onedim_index}. Such an 1-D array for \texttt{struct lattice} gives much better OpenMP scaling.   
\begin{lstlisting}[language=C,numbers=none,label={code:onedim_index},caption={Accessing as \texttt{lattice[index]} rather than \texttt{lattice[ix][iy][iz]}.},basicstyle=\scriptsize]
int64_t totncells  = nmesh_x * nmesh_y * nmesh_z;
cellarray *lattice  = my_malloc(sizeof(cellarray), totncells);
int64_t *nallocated = my_malloc(sizeof(*nallocated), totncells);
int64_t index = ix*nmesh_y*nmesh_z + iy*nmesh_z + iz;
\end{lstlisting}
Now that we know the total number of cells in the entire domain, we need to allocated memory to store the particles in each cell. However, there is no way to 
know the exact number of particles in each cell without processing the entire data-set; so, we pre-allocate with an estimate of the expected number of particles from the 
volume of each 3-D cell (assuming a random distribution). Then, we can allocate memory for each of the \texttt{X/Y/Z} arrays inside each 3-D cell:
\begin{lstlisting}[language=C,numbers=none,label={code:malloc_cellarray},caption={Pre-allocating memory for the \texttt{X/Y/Z} arrays in \texttt{struct cellarray}.},basicstyle=\scriptsize]
for (int64_t index=0;index<totncells;index++) {
  lattice[index].x = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].y = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].z = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].nelements=0;
  nallocated[index] = expected_n;
}
\end{lstlisting}
Here, \texttt{nallocated} is an array that keeps track of the amount of memory alread allocated for each cell. After this step, all cells have been allocated memory for \texttt{expected\_n} 
particles. Now, we can begin to process the individual particles and assigning them to the 3-D cells (if enough memory has already been allocated to assign the new particle). 
\begin{lstlisting}[language=C,numbers=none,label={code:assign_to_cellarray},caption={Assigning the particles to the \texttt{struct cellarray} in the cell.},basicstyle=\scriptsize]
for (int64_t i=0;i<np;i++)  {
  ix=(int)((x[i]-xmin)*xinv) ;
  iy=(int)((y[i]-ymin)*yinv) ;
  iz=(int)((z[i]-zmin)*zinv) ;

@@

\begin{lstlisting}[language=C,numbers=none,label={code:loop_over_first_cells},caption={Looping over all cells in the first data-set.},basicstyle=\scriptsize]

@@

\begin{lstlisting}[language=C,numbers=none,label={code:finding_3D_index_first},caption={Reconstructing 3-D index for \texttt{first} cell in the first data-set.},basicstyle=\scriptsize]

@@

\begin{lstlisting}[language=C,numbers=none,label={code:looping_over_neighbouring_cells},caption={Looping over all the neighbouring cells and taking care 
of PERIODIC boundary conditions},basicstyle=\scriptsize]

@@

\begin{lstlisting}[language=C,numbers=none,label={code:dereference_second_cell},caption={Dereferencing the pointers for the neighbouring (\texttt{second}) cell under consideration.},basicstyle=\scriptsize]

@@

elements representing the second data-set.

\begin{lstlisting}[language=C,numbers=none,label={code:looping_over_first_particles},caption={Looping over all particles in the \texttt{first} cell and accounting for PERIODIC boundary conditions.},basicstyle=\scriptsize]
for(int64_t i=0;i<first->nelements;i++) {
  DOUBLE x1pos=x1[i];
  DOUBLE y1pos=y1[i];
  DOUBLE z1pos=z1[i];
#ifdef PERIODIC
  x1pos += off_xwrap;
  y1pos += off_ywrap;
  z1pos += off_zwrap;
#endif
\end{lstlisting}

\begin{lstlisting}[language=C,numbers=none,label={code:looping_over_second_particles},caption={AVX intrinsics for looping over all particles in the \texttt{second} cell. PERIODIC boundary conditions have 
already been accounted for in \texttt{x1pos,y1pos,z1pos} variables.},basicstyle=\scriptsize]
const AVX_FLOATS m_x1pos = AVX_SET_FLOAT(x1pos);
const AVX_FLOATS m_y1pos = AVX_SET_FLOAT(y1pos);
const AVX_FLOATS m_z1pos = AVX_SET_FLOAT(z1pos);

int64_t j;
for(j=0;j<=(second->nelements-NVEC);j+=NVEC) {
  const AVX_FLOATS x2pos = AVX_LOAD_FLOATS_UNALIGNED(&x2[j]);
  const AVX_FLOATS y2pos = AVX_LOAD_FLOATS_UNALIGNED(&y2[j]);
  const AVX_FLOATS z2pos = AVX_LOAD_FLOATS_UNALIGNED(&z2[j]);
\end{lstlisting}  
The three codes \xir, \xirppi and \wprp diverge somewhat after this point. For \xir, we need to calculate the full 3-D separation, whereas for \xirppi and \wprp 
the separation is the projected distance. I will discuss further implementations in the following sub-sections dedicated to each code. 
\subsubsection{Pair-counting for \texorpdfstring{\xir}{xi(r)}}\label{section:pair_counting_r}
\subsubsection{Pair-counting for \texorpdfstring{\xirppi}{xi(rp,pi)}}\label{section:pair_counting_rp_pi}
\subsubsection{Pair-counting for \texorpdfstring{\wprp}{wp(rp)}}\label{section:pair_counting_wp}
Since the pair-counting in \wprp always assumes PERIODIC boundary conditions {\em and} only computes an auto-correlation, some more optimizations are possible in the
\wprp calculation. First, each individual cell in the \texttt{struct lattice} elements are sorted based on their \texttt{Z} arrays. The \texttt{X/Y} arrays are 
also re-ordered simultaneously. Once all the \texttt{Z} values inside a cell have been sorted, we can avoid double-counting the pairs by changing the \texttt{iiz loop}
to be:\\
\vspace{0.2in}
\texttt{for(int iiz={\bf 0};iiz<=zbin\_refine\_factor;iiz++)} 
\vspace{0.2in}
\\instead of\\
\vspace{0.2in}
\texttt{for(int iiz={\bf -zbin\_refine\_factor};iiz<=zbin\_refine\_factor;iiz++)} 
\vspace{0.2in}

@@

\subsubsection{API for \texorpdfstring{\xir}{xi(r)}}
The interface for the 3-D correlation function, \xir, is through the \texttt{countpairs} function. Here is 
the corresponding function signature:
\begin{lstlisting}[language=C,numbers=none,label={code:API_DD},basicstyle=\scriptsize,caption={API for the 3-D \xir}]
results_countpairs * countpairs(
const int64_t ND1, const DOUBLE * const X1, const DOUBLE * const Y1, const DOUBLE  * const Z1,
const int64_t ND2, const DOUBLE * const X2, const DOUBLE * const Y2, const DOUBLE  * const Z2,
#ifdef USE_OMP
const int numthreads,
#endif
const int autocorr,
const char *binfile);
\end{lstlisting}

@@

The output from \texttt{countpairs} is contained in \texttt{struct results\_countpairs}. The structure
definition is:
\begin{lstlisting}[language=C,numbers=none,label={code:API_DD_struct},basicstyle=\scriptsize,caption={Structure definition for the output of \xir}]
typedef struct{
  uint64_t *npairs;
  DOUBLE *rupp;
  DOUBLE *rpavg;
  int nbin;
} results_countpairs;

>>>>

\documentclass[12pt,titlepage]{article}

\usepackage{geometry}
\usepackage{amsmath}
\usepackage{aas_macros}
\usepackage{rotating}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue,breaklinks=true]{hyperref}
\usepackage{dirtree}
\usepackage{adjustbox}
\usepackage{bytefield}
\usepackage{chngcntr}
\usepackage{alltt}

\setlength{\textwidth}{6.5in}     %%                                   %%
\setlength{\oddsidemargin}{0in}   %% (It is recommended that you       %%
\setlength{\evensidemargin}{0in}  %%  not change these parameters,     %%
\setlength{\textheight}{8.5in}    %%  at the risk of having your       %%
\setlength{\topmargin}{0in}       %%  proposal dismissed on the basis  %%
\setlength{\headheight}{0in}      %%  of incorrect formatting!!!)      %%
\setlength{\headsep}{0in}         %%                                   %%
\setlength{\footskip}{.5in}       %%                                   %%

@@

%% Code paper - so there are lots of code listings
\lstset{
  language=C,                % choose the language of the code
  basicstyle=\small\ttfamily,      % sets font
  numbers=none,                   % where to put the line-numbers
%%   numberstyle=\small, 
%%   stepnumber=1,                   % the step between two line-numbers.
%%   numbersep=10pt,                  % how far the line-numbers are from the code
  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  tabsize=2,                      % sets default tabsize to 2 spaces
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=true,         % sets if automatic breaks should only happen at whitespace
  captionpos=t,  
  xleftmargin=17pt,
  framexleftmargin=17pt,
  framexrightmargin=17pt,
  framexbottommargin=5pt,
  framextopmargin=5pt,
  showstringspaces=false,
  frame=top,frame=bottom,
  escapeinside={(*@}{@*)},
}
%% Change the listings counter to show Section number + listings number (rather than just the listings number)
\AtBeginDocument{\counterwithin{lstlisting}{section}}

@@

Cache locality and hand-written \href{http://en.wikipedia.org/wiki/Advanced_Vector_Extensions}{AVX} intrinsics are the reasons why the code is very fast. However, that also means that the code is 
not very portable. I have tried my best to ensure that the codes work on Linux and MAC OSX. If it does not work for you, particularly 
if you are on a reasonable Linux install, please \href{mailto:manodeep@gmail.com}{email me}. The API for the codes can be considered 
frozen; I will not change the API without changing the MAJOR release version.

The paper associated with the codes is being (slowly) written by me. I hope to submit the paper to Astronomy \& Computing and release 
the codes simultaneously. Once I submit the paper to arXiv, there will be an actual paper arXiv link \href{valid arXiv link}{here}.

@@

\subsection{Getting the Source}
You can obtain the source in two ways: i) Clone the mercurial repo (\texttt{hg clone} \url{https://bitbucket.org/manodeep/corrfunc/}) 
or ii) Download the tar archive (\href{https://bitbucket.org/manodeep/corrfunc/downloads/corrfunc.1.0.0.tar.gz}{corrfunc.\$MAJOR.0.\$MINOR.tar.gz}) and 
unpack it in the directory where you wish to keep the files (\texttt{tar xvzf}\\ \texttt{corrfunc.\$MAJOR.0.\$MINOR.tar.gz}). Here, \texttt{\$MAJOR}
and \texttt{\$MINOR} refer to the major and minor release versions (current \texttt{\$MAJOR=1}, \texttt{\$MINOR=0}). I will only change 
the \texttt{\$MAJOR} version in the highly unlikely event that the API changes.

\subsection{Directory Structure}
The directory structure for the code looks like this:
\dirtree{%
.1 corrfunc.
.2 paper.
.2 xi\_theory.
.3 benchmarks\DotFill
  \begin{minipage}[t]{8cm}
    IDL scripts to run benchmarks{.} 
  \end{minipage}.
.3 bin\DotFill 
  \begin{minipage}[t]{8cm}
    Will be created to copy executable files when you run `make install'{.} 
  \end{minipage}.
.3 examples\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for example C bindings using the static libraries{.} 
  \end{minipage}.
.3 include\DotFill 
  \begin{minipage}[t]{8cm}
    Header files for static libraries{.} 
  \end{minipage}.
.3 io\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for reading in data{.} 
  \end{minipage}.
.3 lib\DotFill 
  \begin{minipage}[t]{8cm}
    Will be created to copy static libraries and python library after you run `make'{.} 
  \end{minipage}.
.3 python\_bindings\DotFill 
  \begin{minipage}[t]{8cm}
    Source files to generate python bindings{.} 
  \end{minipage}.
.3 tests\DotFill 
  \begin{minipage}[t]{8cm}
    Correct outputs for tests{.}
  \end{minipage}.
.4 data\DotFill 
  \begin{minipage}[t]{8cm}
    Mock galaxy catalogs for tests{.} 
  \end{minipage}.
.3 utils\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for creating 3-D grid and helper routines{.} 
  \end{minipage}.
.3 xi\_of\_r\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $\xi(r)${.} 
  \end{minipage}.
.3 xi\_rp\_pi\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $\xi(r_p,\pi)${.} 
  \end{minipage}.
.3 wp\DotFill 
  \begin{minipage}[t]{8cm}
    Source files for $w_p(r_p)${.} 
  \end{minipage}.
}

\subsection{Compilation Options}
There are a few code options that control both the Science case and the code compilation. All of these 
options are located in `common.mk' in the base directory (`corrfunc'). Edit the first few lines to set these 
options (see Table.~\ref{table:options} for details) :
\begin{itemize}
\item Science options -- PERIODIC, OUTPUT\_RPAVG
\item Code options -- DOUBLE\_PREC, USE\_AVX and USE\_OMP
\end{itemize}

@@

\flushleft The \texttt{fast-food} file consists of a header:
\begin{lstlisting}[caption={The header format for \texttt{fast-food} files}]
int idat[5];
float fdat[9];
float znow;
\end{lstlisting}
For the purposes of these correlation function codes, the only useful quantity is \texttt{idat[1]} which contains \texttt{N} -- the number of particles 
in the data file.

@@

\stdsection{Code Design}
The following sections are taken directly from the associated \href{valid arXiv link}{paper}. The fundamental idea 
behind the code can be broken down into the following steps:
\begin{itemize}
\item Given the max. separation, \rmax, for the correlation function, grid the entire the particle distribution with cell width \rmax. In case of 
\xirppi and \wprp, the \texttt{X/Y} bins are \rmax while the \texttt{Z} bins is \pimax.
\item In each 3-D cell, store all the particles that are located in that cell in contiguous \texttt{X/Y/Z} arrays (see Listing~\ref{code:assign_to_cellarray}). The particle data are contained in a structure 
defined in Listing~\ref{code:celldefn}.
\item Loop over all particles in every cell (this loop uses OpenMP parallelization if USE\_OMP is enabled). This is the \texttt{index1} loop in 
\texttt{countpairs.c, countpairs\_rp\_pi.c and countpairs\_wp.c}. The loop variable for the loop over each particle is \texttt{i}.
\item Now, for every particle in the input list, we only need to look at particles in the neighbouring cells (see Fig.~\ref{fig:grid}). 
\item Loop over neighbouring cells in all 3 dimensions. Corresponding variables are \texttt{iiix, iiiy, iiiz}. These variables are constructed from the 
loop variables \texttt{iix, iiy, iiz} to account for periodic boundary conditions. \footnote{I apologize for this atrocious variable naming scheme.}
\item Once we have a triplet of \texttt{iiix, iiiy, iiiz}, we can construct the index for the neighbouring cell. Variable \texttt{index2} gives the index 
for the neighbouring cell. 
\item Compute the distances between target particle and some particle bunch (where a chunk is 4 double or 8 floats) using AVX intrinsics. The slower, non-AVX 
version of the code computes the distances serially and does not take advantage of the CPU architecture. This is the \texttt{j} loop in 
\texttt{countpairs.c, countpairs\_rp\_pi.c, countpairs\_wp.c}. 
\item Check if any of the distances are less than \rmax. If not, continue processing the data. If yes, then update the histograms using 
AVX \texttt{bit-masks} and \texttt{popcount}. 
\item Continue until all particles in neighbouring cell are done. 
\end{itemize}

@@

\subsection{How to Maintain Cache Locality within the Grid}
For all pairs around a given target galaxy, we need to compute distances to all points within all neighbouring 3-d cells.
We ensure that the particle locations are contiguous by moving them into the following \texttt{C struct} in the order in which they arrive.
\begin{lstlisting}[label={code:celldefn},caption={Definition of the cellarray structure. This 
structure contains the \texttt{X/Y/Z} positions of all the particles that are in one 3-D cell. }]
typedef struct{
  DOUBLE *x;
  DOUBLE *y;
  DOUBLE *z;
  int64_t nelements;
} cellarray;
\end{lstlisting}
The code \texttt{gridlink.c} takes in an input list of 3 arrays \texttt{X/Y/Z} and grids them into a regular 3-D grid using the specified bins parameter
\texttt{max\_x\_size} or \texttt{max\_y\_size} or \texttt{max\_z\_size} for the \texttt{X/Y/Z} axes respectively. \footnote{In practice, the bins may be further 
subdivided using the corresponding bin refine factors. These bin refine factors seem to influence runtime the most, you should experiment with a few values of 
\texttt{bin\_refine\_factor} and \texttt{zbin\_refine\_factor} to see what produces the best runtimes for your typical scenario.}
Once the number of grid cells along each axes has been determined, we allocate memory for the \texttt{struct lattice}. This \texttt{struct lattice} is a 
declared as an 1-D array; the conversion from the three indices in 3-D (\texttt{ix,iy,iz}) to a single index (\texttt{index}) happens through the last 
line in Listing~\ref{code:onedim_index}. Such an 1-D array for \texttt{struct lattice} gives much better OpenMP scaling.   
\begin{lstlisting}[label={code:onedim_index},caption={Accessing as \texttt{lattice[index]} rather than \texttt{lattice[ix][iy][iz]}.}]
int64_t totncells  = nmesh_x * nmesh_y * nmesh_z;
cellarray *lattice  = my_malloc(sizeof(cellarray), totncells);
int64_t *nallocated = my_malloc(sizeof(*nallocated), totncells);
int64_t index = ix*nmesh_y*nmesh_z + iy*nmesh_z + iz;
\end{lstlisting}
Now that we know the total number of cells in the entire domain, we need to allocated memory to store the particles in each cell. However, there is no way to 
know the exact number of particles in each cell without processing the entire data-set; so, we pre-allocate with an estimate of the expected number of particles from the 
volume of each 3-D cell (assuming a random distribution). Then, we can allocate memory for each of the \texttt{X/Y/Z} arrays inside each 3-D cell:
\begin{lstlisting}[label={code:malloc_cellarray},caption={Pre-allocating memory for the \texttt{X/Y/Z} arrays in \texttt{struct cellarray}.}]
for (int64_t index=0;index<totncells;index++) {
  lattice[index].x = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].y = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].z = my_malloc(sizeof(DOUBLE),expected_n);
  lattice[index].nelements=0;
  nallocated[index] = expected_n;
}
\end{lstlisting}
Here, \texttt{nallocated} is an array that keeps track of the amount of memory alread allocated for each cell. After this step, all cells have been allocated memory for \texttt{expected\_n} 
particles. Now, we can begin to process the individual particles and assigning them to the 3-D cells (if enough memory has already been allocated to assign the new particle). 
\begin{lstlisting}[label={code:assign_to_cellarray},caption={Assigning the particles to the \texttt{struct cellarray} in the cell.}]
for (int64_t i=0;i<np;i++)  {
  ix=(int)((x[i]-xmin)*xinv) ;
  iy=(int)((y[i]-ymin)*yinv) ;
  iz=(int)((z[i]-zmin)*zinv) ;

@@

\begin{lstlisting}[label={code:loop_over_first_cells},caption={Looping over all cells in the first data-set.}]

@@

\begin{lstlisting}[label={code:finding_3D_index_first},caption={Reconstructing 3-D index for \texttt{first} cell in the first data-set.}]

@@

\begin{lstlisting}[label={code:looping_over_neighbouring_cells},caption={Looping over all the neighbouring cells and taking care of PERIODIC boundary conditions.}]

@@

\begin{lstlisting}[label={code:dereference_second_cell},caption={Dereferencing the pointers for the neighbouring (\texttt{second}) cell under consideration.}]

@@

elements representing the second data-set. Now, we have to compute all possible pair-wise separations between these two data-sets. We begin with a loop over the elements in 
\texttt{first}:
\begin{lstlisting}[label={code:looping_over_first_particles},caption={Looping over all particles in the \texttt{first} cell and accounting for PERIODIC boundary conditions.}]
for(int64_t i=0;i<first->nelements;i++) {
  DOUBLE x1pos=x1[i];
  DOUBLE y1pos=y1[i];
  DOUBLE z1pos=z1[i];
  x1pos += off_xwrap;
  y1pos += off_ywrap;
  z1pos += off_zwrap;
If PERIODIC boundary conditions are enabled, then \texttt{off\_xwrap}, \texttt{off\_ywrap} and \texttt{off\_zwrap} have been declared and initialized in Listing~\ref{code:looping_over_neighbouring_cells}.
By wrapping the elements if the first data-set, we can avoid these operations in the following \texttt{j-loop} over all particles in \texttt{second}.

\begin{lstlisting}[label={code:looping_over_second_particles},caption={AVX intrinsics for looping over all particles in the \texttt{second} cell. PERIODIC boundary conditions have 
already been accounted for in \texttt{x1pos,y1pos,z1pos} variables.}]
const AVX_FLOATS m_x1pos = AVX_SET_FLOAT(x1pos);
const AVX_FLOATS m_y1pos = AVX_SET_FLOAT(y1pos);
const AVX_FLOATS m_z1pos = AVX_SET_FLOAT(z1pos);

int64_t j;
for(j=0;j<=(second->nelements-NVEC);j+=NVEC) {
  const AVX_FLOATS x2pos = AVX_LOAD_FLOATS_UNALIGNED(&x2[j]);
  const AVX_FLOATS y2pos = AVX_LOAD_FLOATS_UNALIGNED(&y2[j]);
  const AVX_FLOATS z2pos = AVX_LOAD_FLOATS_UNALIGNED(&z2[j]);

  const AVX_FLOATS m_xdiff = AVX_SUBTRACT_FLOATS(m_x1pos,x2pos);
  const AVX_FLOATS m_ydiff = AVX_SUBTRACT_FLOATS(m_y1pos,y2pos);
  const AVX_FLOATS m_zdiff = AVX_SUBTRACT_FLOATS(m_z1pos,z2pos);
\end{lstlisting}

The three codes \xir, \xirppi and \wprp diverge somewhat after this point. For \xir, we need to calculate the full 3-D separation, whereas for \xirppi and \wprp 
the separation is the projected distance. I will discuss further implementations in the following sub-sections dedicated to each code. 
\subsubsection{Pair-counting for \texorpdfstring{\xir}{xi(r)}}\label{section:pair_counting_r}
Pair-counting for \xir is the most straight-forward. Once all the \texttt{NVEC} separations have been computed, we use bit-masks to check if any separations fall 
within the range \texttt{sqr\_rpmin} and \texttt{sqr\_rpmax}. If not, we \texttt{continue} with the \texttt{j-loop}. 
\begin{lstlisting}[label={code:masks_xi},caption={Bit-masks in \xir.}]
m_mask_left = AVX_COMPARE_FLOATS(r2,m_sqr_rpmax,_CMP_LT_OS);
if(AVX_TEST_COMPARISON(m_mask_left) == 0) {
  continue;
}

const AVX_FLOATS m_mask = AVX_BITWISE_AND(m_mask_left, AVX_COMPARE_FLOATS(r2, m_sqr_rpmin, _CMP_GE_OS));
if(AVX_TEST_COMPARISON(m_mask) == 0) {
  continue;
}
r2 = AVX_BLEND_FLOATS_WITH_MASK(m_sqr_rpmax, r2, m_mask);
m_mask_left = AVX_COMPARE_FLOATS(r2,m_sqr_rpmax,_CMP_LT_OS);
\end{lstlisting}

\subsubsection{Pair-counting for \texorpdfstring{\xirppi}{xi(rp,pi)}}\label{section:pair_counting_rp_pi}

\subsubsection{Pair-counting for \texorpdfstring{\wprp}{wp(rp)}}\label{section:pair_counting_wp}
Since the pair-counting in \wprp always assumes PERIODIC boundary conditions {\em and} only computes an auto-correlation, some more optimizations are possible in the
\wprp calculation. First, each individual cell in the \texttt{struct lattice} elements are sorted based on their \texttt{Z} arrays. The \texttt{X/Y} arrays are 
also re-ordered simultaneously. Once all the \texttt{Z} values inside a cell have been sorted, we can avoid double-counting the pairs by changing the \texttt{iiz loop}
to be:\\
\vspace{0.2in}
\texttt{for(int iiz={\bf 0};iiz<=zbin\_refine\_factor;iiz++)} 
\vspace{0.2in}
\\instead of\\
\vspace{0.2in}
\texttt{for(int iiz={\bf -zbin\_refine\_factor};iiz<=zbin\_refine\_factor;iiz++)} 
\vspace{0.2in}

@@

\subsubsection{API for \texorpdfstring{\xir}{xi(r)}}
The interface for the 3-D correlation function, \xir, is through the \texttt{countpairs} function. Here is 
the corresponding function signature:
\begin{lstlisting}[label={code:API_DD},caption={API for the 3-D \xir.}]
results_countpairs * countpairs(
const int64_t ND1, const DOUBLE * const X1, const DOUBLE * const Y1, const DOUBLE  * const Z1,
const int64_t ND2, const DOUBLE * const X2, const DOUBLE * const Y2, const DOUBLE  * const Z2,
#ifdef USE_OMP
const int numthreads,
#endif
const int autocorr,
const char *binfile);
\end{lstlisting}

@@

The output from \texttt{countpairs} is contained in \texttt{struct results\_countpairs}. The structure
definition is:
\begin{lstlisting}[label={code:API_DD_struct},caption={Structure definition for the output of \xir.}]
typedef struct{
  uint64_t *npairs;
  DOUBLE *rupp;
  DOUBLE *rpavg;
  int nbin;
} results_countpairs;
